---
output:
  html_document: default
  pdf_document: default
  word_document: default
---
#Analysis of White Wine Quality


###Mylu Praveen

========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Load all of the packages that you end up using in your analysis in this code
# chunk.
suppressMessages(library(reshape))
suppressMessages(library(reshape2))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(maps))
suppressMessages(library(RColorBrewer))
suppressMessages(library(GGally))
suppressMessages(library(scales))
suppressMessages(library(memisc))
suppressMessages(library(gridExtra))


library(ggplot2)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv("wineQualityWhites.csv")
```


#Introduction:

This tidy data set contains 4,898 white wines with 11 variables on quantifying
the chemical properties of each wine. At least 3 wine experts rated the quality
of each wine, providing a rating between 0 (very bad) and 10 (very excellent).


#Variables contained in the Data set:

```{r echo=FALSE, Checking_variables}
str(wine)
```


#Data Munging:

We need to remove variable X as it represents serial number in the data.


```{r echo=FALSE, Removing_X}
wine$X <- NULL
str(wine)
```


#Description of Attributes from Data:


1)  fixed acidity: 
    most acids involved with wine or fixed or nonvolatile
    (do not evaporate readily).

2)  volatile acidity:
    The amount of acetic acid in wine, which at too high of levels can lead to
    an unpleasant, vinegar taste.

3)  citric acid: found in small quantities, citric acid can add 'freshness' 
    and flavor to wines.

4)  residual sugar: the amount of sugar remaining after fermentation stops, it's
    rare to find wines with less than 1 gram/liter and wines with greater than 
    45 grams/liter are considered sweet.

5)  chlorides: the amount of salt in the wine.

6)  free sulfur dioxide: the free form of SO2 exists in equilibrium between 
    molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial
    growth and the oxidation of wine.

7)  total sulfur dioxide: amount of free and bound forms of S02; in low 
    concentrations, SO2 is mostly undetectable in wine, but at free SO2 
    concentrations over 50 ppm, SO2 becomes evident in the nose and taste of 
    wine.

8)  density: the density of water is close to that of water depending on the 
    percent alcohol and sugar content.

9)  pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
    to 14 (very basic); most wines are between 3-4 on the pH scale.

10) sulphates: a wine additive which can contribute to sulfur dioxide gas (S02) 
    levels, wich acts as an antimicrobial and antioxidant.

11) alcohol: the percent alcohol content of the wine

    Output variable (based on sensory data):
    
12) quality (score between 0 and 10).



# New Variables:

1) Volatile.type: represents taste where low level tends to have a good taste.
 
2) citric.type: represents amount of flavour in wine.

3) quality.type: represents quality on a basis of low, medium and high. 

4) total.acidity: represents total amount of acidity in wine. 
   (total.acidity = fixed.acidity + volatile.acidity)
 

# Adding New variables:

```{r echo = FALSE, Adding_variables}
wine["volatile.type"] <- wine$volatile.acidity
wine$volatile.type <- cut(wine$volatile.acidity, breaks = c(-Inf, 0.5, 
                            0.6, +Inf), labels = c("low", "middle", "high"))


wine["citric.type"] <- wine$citric.acid
wine$citric.type <- cut(wine$citric.acid, breaks = c(-Inf, 0.5, 1, +Inf), 
                            labels = c("less", "medium", "more"))

wine["total.acidity"] <- wine$fixed.acidity + wine$volatile.acidity


wine["quality.factor"] <- factor(wine$quality, ordered = TRUE)

```

#Summary of Data:


```{r echo=FALSE}
summary(wine)
```


# Univariate Plots Section


```{r echo=FALSE, Univariate_Plot1}
create_plot <- function(varname, binwidth = 0.05) {
  return(ggplot(aes_string(x = varname), data = wine) + 
           geom_histogram(binwidth = 0.1, bins = 1000))
}
create_plot('fixed.acidity', 100) + xlim(0, 15)
```

* fixed.acidity has an normal distribution with a mean of 6.855. It has a median
  of value 6.8 and there are very few outliers after the fixed.acidity value of 
  10 in this plot . The maximum count for fixed.acidity is nearly 310. 
  fixed.acidity has an interval range of {3.80, 14.2}.


```{r echo=FALSE, Univariate_Plot2}
create_plot('volatile.acidity', 100) + xlim(0,1.5)
```

* volatile.acidity is a positive skew distribution with a mean of 0.08. It has
  a median of a value 0.26 and there are no outliers in this plot. The maximum 
  count for volatile.acidity is almost 2000. volatile.acidity has an interval 
  range of {0.08, 1.10}.
  

```{r echo=FALSE, Univariate_Plot3}
create_plot('citric.acid', 100) + xlim(0, 2)
```

* citric.acid is a normal distribution with a mean of 0.3342. It has
  a median of a value 0.32 and there are no outliers in this plot. The maximum 
  count for volatile.acidity is almost 2250. citric.acid has an interval 
  range of {0, 1.66}.
  
  
```{r echo=FALSE, Univariate_Plot4}
create_plot('residual.sugar', 100) + xlim(0,70)
```

* residual.sugar is a positive skew distribution with a mean of 6.391. It has
  a median of a value 5.2 and there are few outliers in this plot occuring after
  the residual.sugar value of 20. The maximum count for residual.sugar is 
  almost 185. residual.sugar has an interval range of {0.60, 65.8}.
  
  
```{r echo=FALSE, Univariate_Plot5}
create_plot('chlorides', 100) + xlim(0,0.35)
```

*  chlorides is a positive skew distribution with a mean of 0.045. It has
  a median of a value 0.043 and there are no outliers in this plot. The maximum 
  count for chlorides is almost 1100. chlorides has an interval 
  range of {0.009, 0.346}.
  

```{r echo=FALSE, Univariate_Plot6}
create_plot('total.sulfur.dioxide', 100) + xlim(0, 450)
```


* total.sulfur.dioxide is a bimodal distribution with a mean of 138.4. It has
  a median of a value 134 and there are no outliers in this plot. The maximum 
  count for total.sulfur.dioxide is nearly 50. total.sulfur.dioxide has an 
  interval range of {9, 440}.
  
  
```{r echo=FALSE, Univariate_Plot7}
create_plot('pH', 100) + xlim(0,4)
```


* pH is a binomial distribution with a mean of 3.188. It has
  a median of a value 3.18 and there are no outliers in this plot. The maximum 
  count for pH is almost 1300. pH has an interval 
  range of {2.72, 3.82}.
  
  
```{r echo=FALSE, Univariate_Plot8}
create_plot('sulphates', 100) + xlim(0, 1.1)
```


* sulphates is a binomial distribution with a mean of 0.4898. It has
  a median of a value 0.47 and there are no outliers in this plot. The maximum 
  count for sulphates is almost 1700. sulphates has an interval 
  range of {0.22, 1.08}.
  

```{r echo=FALSE, Univariate_Plot9}
create_plot('alcohol', 100) + xlim(0,15)
```

* alcohol is a positive skew distribution with a mean of 10.51. It has
  a median of a value 10.4 and there is an single outlier near to 14 in this 
  plot. The maximum count for alcohol is almost 235. alcohol has an 
  interval range of {8, 14.2}.


```{r echo=FALSE, Univariate_Plot10}
ggplot(aes(x = as.factor(quality)), data = wine) + geom_bar() 
```

* quality is a bimodal distribution with a mean of 5.878. quality is an 
  ordered categorical variableIt has a median of a value 6 and there are no 
  outliers in this plot. The maximum count for volatile.acidity is almost 2250. 
  quality has an interval range of {0.08, 1.10}.


# New variables Plots Section

```{r echo=FALSE, new_variables_plot }
p11 = ggplot(aes(x = volatile.type), data = wine) + geom_bar()
p12 = ggplot(aes(x = citric.type), data = wine) + geom_bar()
p13 = ggplot(aes(x = total.acidity), data = wine) + geom_histogram(bins = 30)
p14 = ggplot(aes(x = quality.factor), data = wine) + geom_bar()
grid.arrange(p11,p12,p13,p14,ncol = 2)
```


* White wine produces more amount of low volatile.type which says it consists 
  of more amount of good taste. low volatile.type has a count of 4742, 
  middle volatile.type of 90 and high volatile.type of 66.
  
* White wine consists of less amount of flavour in wine. less citric.type has
  a count of 4584, medium citric.type of 312 and more citric.type of 2.
  
* White wine produces a better quality wine. normal quality.type has a count of
  20, better quality.type of 3818 and good quality.type of 1060.

* total amount of acidity is an binomial distribution with a mean of 7.133. It
  has a median of a value 7.07 and there are no outliers in this plot. The 
  maximum count for total.acidity is almost 900. total.acidity has an interval 
  range of {4.11, 14.47}.


# Transforming Variables 

```{r echo=FALSE, transformed_univariate_Plots}
p11 <- ggplot(aes(x = volatile.acidity), data = wine) + 
                          geom_histogram(binwidth = 0.05, bins = 30) + 
        scale_x_log10()
p12 <- ggplot(aes(x = residual.sugar), data = wine) +
                          geom_histogram(binwidth = 0.05, bins = 30) + 
       scale_x_log10(breaks = scales::trans_breaks("log10", function(x) 10^x))
p13 <- ggplot(aes(x = chlorides), data = wine) +
                         geom_histogram(binwidth = 0.05, bins = 30) +
      scale_x_log10(breaks = scales::trans_breaks("log10", function(x) 10^x))
p14 <- ggplot(aes(x = alcohol), data = wine) + 
                         geom_histogram(binwidth = 0.05, bins = 30) + 
       scale_x_log10(breaks = scales::trans_breaks("log10", function(x) 10^x))
grid.arrange(p11,p12,p13,p14,ncol = 2)
```

  we have used breaks parameter, so that we can observe sufficient tickmarks
  in every plot and find the values that are reflected in a particular position
  in the log scaled axis.

* volatile.acidity has an binomial distribution after applying transformation.

* residual.sugar has an bimodal distribution. It has an interval ranging from
  {1.000000, 31.622777}.

* chlorides has an binomial distribution. It has an interval ranging from
  {0.01000000, 0.31622777}

* alcohol has also an binomial distribution.  It has an interval ranging from
  {7.943282, 15.848932}


# Correlation between variables

1) quality vs alcohol

```{r echo= FALSE}
cor.test(wine$quality, wine$alcohol)
```

* The correlation between quality and alcohol is 0.4355747. 

2) alcohol vs density

```{r echo= FALSE}
cor.test(wine$alcohol, wine$density)
```

* The correlation between alcohol and density is -0.7801376.

3) chlorides vs sulphates

```{r echo= FALSE}
cor.test(wine$chlorides, wine$sulphates)
```

* The correlation between chlorides and sulphates is 0.01676288.

4) chlorides vs density

```{r echo= FALSE}
cor.test(wine$chlorides, wine$density)
```

* The correlation between chlorides and density is 0.2572113 .

5) alcohol vs density

```{r echo= FALSE}
cor.test(wine$alcohol, wine$density)
```

* The correlation between alcohol and density is -0.7801376.


# Bivariate Plots Section

### Volatile.acidity vs citric.acid

```{r echo= FALSE}
cor.test(wine$volatile.acidity, wine$citric.acid)
ggplot(aes(x = volatile.acidity, y = citric.acid), data = wine) + 
               geom_point(alpha = 1/10, position = position_jitter(h = 0))
```

* From volatile.acidity vs citric.acid plot, we observe a negative correlation 
between them. The data lies mostly between the 2nd and 3rd quartile. 
volatile.acidity depends on the taste of the wine where less amount of volatile
gives the better taste and citric.acid depends on the addition of flavour and
freshness to wine where higher the amount of citric acid increases the freshness
and flavour to the wine. Therefore I conclude that the wines that we can prefer
containing a volatile.acidity of nearly 0.18 which would be of much taste, also
containing the citric.acid of 1.7. some amount of outliers also exists in this 
plot.


### free.sulfur.dioxide vs total.sulfur.dioxide

```{r echo= FALSE}
cor.test(wine$free.sulfur.dioxide, wine$total.sulfur.dioxide)
ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide), data = wine) +
              geom_point(alpha = 1/10, position = position_jitter(h = 0))
```


* From free.sulfur.dioxide vs total.sulfur.dioxide plot, we observe a positive
correlation between them. The data mostly lies in the interquartile range 
i.e. between 1st and 3rd quantile.


# Multivariate Plots Section

### density vs pH colored by quality.factor

```{r echo=FALSE}
cor.test(wine$density, wine$pH)
ggplot(aes(x = density, y = pH), data = wine, color = factor(quality)) +
  geom_point(alpha = 1/10, position = position_jitter(h = 0))  + 
  facet_wrap(~quality.factor) + geom_smooth(method = "lm", se = FALSE, size = 1) 
     
```


* From this density vs pH plot, we used facet for quality.factor and also found 
  negative correlation between density and pH. Later on we used geom_smooth 
  to add a summary line in the plot.


### alcohol vs sulphates facet by quality

```{r echo=FALSE}
cor.test(wine$alcohol, wine$sulphates)
ggplot(aes(x = alcohol, y = sulphates), data = wine) + 
  geom_point(alpha = 1/10, position = position_jitter(h = 0))  + 
                    facet_wrap(~quality) 
```


* From this plot we observe a negative correlation between alcohol and sulphates
  where facet was done by quality and found quality of wine is better for 
  alcohol and sulphates.   


# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
qplot(as.numeric(as.character(quality)), data = wine, binwidth = 1, 
               main = "Quality of wine")  + 
  scale_x_continuous(breaks = seq(2,10,1), limits = c(2,10)) + 
  scale_y_sqrt(breaks = seq(0,5000,500)) + xlab("quality")
         
                     
```

### Description One

I have choosen this quality plot as it represents most of my finding in this 
analysis, most of the quality levels occur in 5th and 6th level and level 9 
seems to be very low quality. quality seems to be the most important feature and
property analyzing this data.


### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = quality, y = alcohol), data = wine) + 
  geom_boxplot(aes(color = quality.factor)) + xlab("quality") +
  ylab("alcohol (%) ") + ggtitle("quality vs alcohol colored by quality.factor") 
         
```



### Description Two

This plot shows an relationship between quality and alcohol where we can 
observe that as alcohol percent increases in wine increases the quality of the 
wine. The reason i choosed this boxplot to observe higher alcohol percentage 
based on the quality. In this plot i have colored quality.factor which will 
represent all quality types with individual color.
 
 
### plot Three

```{r echo=FALSE, Plot_three}
ggplot(aes(x = density, y = alcohol), data = wine) + geom_point(alpha = 1/10, 
        position = position_jitter(h = 0)) +  facet_wrap(~quality.factor) +
 scale_x_continuous(limits = c(0.9,1.1)) +  
      scale_y_continuous(limits = c(8,15)) + xlab("density") +ylab("alcohol") +
          ggtitle("density vs alcohol facet by quality.factor")
            
         
```


### Description Three

From the above plot, we observe that quality 7 has the highest alcohol 
percentage and quality 5 has the lowest alcohol percentage.
we have a correlation of -0.78 between density and alcohol.
The reason i choosed this plot is to find the alcohol percentage based
on the quality. I have used facet_wrap command to analyze density vs alcohol in
each quality and also adjusted axis limit inorder to reduce overplotting.


# Reflection

From the given white wine quality data set i have analyzed some the data with 
the help of some major variables. I have plotted necessary correlated plots of 
various variables. 

* Density will increase as alcohol percentage decreases.

* volatile.acidity and citric.acid are negatively correlated where 
  volatile.acidity depends on the taste of the wine where less amount of 
  volatilegives the better taste and citric.acid depends on the addition of 
  flavour and freshness to wine where higher the amount of citric acid 
  increases the freshness and flavour to the wine.

* free.sulfur.dioxide and total.sulfur.dioxide are positively correlated.

* density and pH are negatively correlated.

* alcohol and sulphates are negatively correlated.

* In quality vs alcohol, quality 7 has the highest alcohol percentage and 
  quality 5 has the lowest alcohol percentage. where they have a correlation of 
  -0.78 between density and alcohol.

* alcohol percent increases in wine also increases the quality of the wine in
  good quality type.


The intresting feature in analyzing is by finding the final plots with each plot
with different type of plot such as boxplot and scatterplot. I have found and 
anlayzed different types of plot in univariate, Bivariate and Multivariate
plots. 

There is no much difficulty in analyzing the plots where it would be even more 
better if more properties of wine are included in the data set.

My future exploration from this data will be adding some other additional 
properties such as age of the wine, type of the wine etc. It would help us to 
analyze more information and create better models.



